---
title: "Managing metadata schema"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Manage Schema}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

# Setup
```{r setup, eval=FALSE}
## Install package
devtools::document("~/OmicsMLRepo/OmicsMLRepoCuration")
install.packages("~/OmicsMLRepo/OmicsMLRepoCuration", repos = NULL, type="source")
```

```{r}
library(OmicsMLRepoCuration)
```

```{r}
## Setup directory
schema_dir <- "~/OmicsMLRepo/OmicsMLRepoCuration/inst/schema"
```


# Compare official vs. local versions
```{r}
## Cloud copy (i.e., my local work)
url <- "https://docs.google.com/spreadsheets/d/1KK-rayNsxcYG45ToBPT5_HWLxoOpbCLZlbbZ-Y72fRc/edit?usp=sharing"
ss <- googledrive::as_id(url)
dd <- googlesheets4::read_sheet(ss = ss, sheet = "cmd")

## Official copy
cmd <- readr::read_csv("~/Projects/cMD_architecture/curatedMetagenomicDataCuration/inst/extdata/cmd_data_dictionary.csv")
cmd <- cmd[order(cmd$col.name),]
```


```{r}
print(paste("Are they identical?:", identical(dd, cmd)))

library(diffdf)
diffdf(dd, cmd) # 'BASE' is the local work (my version in Google Sheet), 'COMP' is the official version in cMDC
```


# Sync
## Local (Google Sheet) --> Official (cMDCuration GH)
If your local work is done in Google Sheet
```{r cloud_to_local, eval=FALSE}
write.csv(dd, file.path(schema_dir, "cmd_data_dictionary.csv"), row.names = FALSE)
write.csv(dd, "~/Microbiome/curatedMetagenomicDataCuration/inst/extdata/cmd_data_dictionary.csv", row.names = FALSE)
```

```{r}
## csv to yaml
tableToYamlSchema(dd, 
                     output_file = file.path(schema_dir, "cmd_data_dictionary.yaml"))

## csv to linkml
tableToLinkmlSchema(dd,
                       schema_version = "0.0.11",
                       output_file = file.path(schema_dir, "cmd_data_dictionary_linkml.yaml"))
```

The LinkML output is ontology-grounding-ready, so downstream tools (e.g. the
`metacurator` grounding engine) can bind fields to their ontology branches
directly. Two data-dictionary columns drive this:

- **`static.enum`** — the pipe-separated ontology term ids are written as LinkML
  `meaning` CURIEs on each permissible value. For example `body_site`
  (`allowedvalues = feces|oral cavity|...`, `static.enum = UBERON:0001988|...`)
  becomes an enum whose `feces` value carries `meaning: UBERON:0001988`.
- **`dynamic.enum`** — the ontology root(s) are written as a LinkML
  `reachable_from` dynamic enum (one `include: reachable_from` per root, so
  multi-ontology roots union). `dynamic.enum.property = children` yields direct
  subclasses (`is_direct: true`); anything else yields transitive descendants.
  For example `disease` (`dynamic.enum = NCIT:C7057;EFO:0000408`) binds to both
  the NCIT and EFO branches.

Fields whose `static.enum` id has no matching label (e.g. a bare id on a
`dynamic_enum;static_enum` field) cannot be expressed as a permissible value and
are omitted; give the value a label in `allowedvalues` to represent it.

```{r echo=FALSE, eval=FALSE}
## Local (Google Sheet) to GitHub sync.
syncSchema(schema_dir = "~/OmicsMLRepo/OmicsMLRepoCuration/inst/schema",
           schema_version = "0.0.12")
```






## Official (cMDCuration GH) --> Local (Google Sheet) 
If your local work (e.g., Google Sheet) is behind the official version


```{r gh_to_cloud, echo=FALSE, eval=FALSE}
# ## Sync the official GitHub version to Google Drive
# googlesheets4::write_sheet(cmd, ss = ss, sheet = "cMD_data_dictionary")
```

```{r load_yaml_schema}
schema_file <- system.file("schema", "cmd_data_dictionary.yaml", package = "OmicsMLRepoCuration")
schema <- loadMetadataSchema(schema_file)
dd <- exportSchemaToCsv(schema)
```
