---
title: "Choosing Representative Nodes for a Set of Ontology Terms"
author: "Kaelyn Long"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{representativenodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include = FALSE}
## Set knitr hook to limit output lines
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold",
  results = "hold",
  output.lines = 10,
  fig.align = "center",
  fig.width = 8,
  fig.height = 8
)

options(knitr.kable.NA = "",
        rmarkdown.html_vignette.check_title = FALSE)
```

```{r package_setup, include = FALSE}
library(OmicsMLRepoCuration)
library(readr)
library(stringr)
library(dplyr)
library(jsonlite)
library(igraph)
library(kableExtra)
```

# Representative node purpose

The `mapNodes()` function and accompanying accessory functions were created to
assist curators in choosing a small number of ontology terms as "representative
nodes" to include in a dynamic enum function.

These functions make use of the `igraph` package to create network
representations of the ontology sections containing input ontology terms. The
`igraph` package also has a fast clustering method that is used to break down
large network graphs into clusters of similar nodes.

These functions also incorporate a "cutoff" value that indicates when to exclude
"high-traffic" terms as representative node candidates. The cutoff indicates a
percentage of the graph and if a single term is an ancestor of that percentage
of the graph or more, it will be designated as "high-traffic."

---

# Basic implementation

The basic implementation of `mapNodes()` is very straightforward. The function
takes a character vector of ontology term ids and an optional value for the
cutoff. The default value is 0.25.

```{r basic_imp}
dir <- system.file("extdata", package = "OmicsMLRepoCuration")
ids <- read_lines(file.path(dir, "sample_ids.csv"))
t <- mapNodes(ids)
```
```{r display_basic_imp, results = "asis", echo = FALSE}
t %>%
    mutate(original_covered = str_trunc(original_covered, 20, "right")) %>%
    kbl(col.names = c("Ontology Term", "Ontology Term ID",
                      "Original Term Ontology", "Original Covered",
                      "Num. Original Covered", "Num. Original"),
        align = "c",
        caption = "Chosen representative nodes with default cutoff") %>%
    kable_styling(bootstrap_options = c("striped",
                                        "hover",
                                        "condensed",
                                        "responsive"),
                  font_size = 10)
```

<br>

Here we can see the difference in results at different cutoff values.

```{r cutoffs_imp}
nodes0.05 <- mapNodes(ids, cutoff = 0.05)
nodes0.25 <- mapNodes(ids, cutoff = 0.25)
nodes0.95 <- mapNodes(ids, cutoff = 0.95)
```

```{r display_cutoffs_imp, results = "asis", echo = FALSE}
ml <- max(nrow(nodes0.05), nrow(nodes0.25), nrow(nodes0.95))

t05 <- nodes0.05$ontology_term
t25 <- nodes0.25$ontology_term
t95 <- nodes0.95$ontology_term

length(t05) <- ml
length(t25) <- ml
length(t95) <- ml

c <- data.frame("cutoff05" = t05,
                "cutoff25" = t25,
                "cutoff95" = t95)

c %>%
    kbl(col.names = c("Cutoff = 0.05", "Cutoff = 0.25", "Cutoff = 0.95"),
        align = "c",
        caption = "Differences in number of nodes chosen at different cutoff values") %>%
    kable_styling(bootstrap_options = c("hover",
                                        "condensed",
                                        "responsive"),
                  font_size = 10)
```

---

# Function walkthrough

Now we will walk through the main wrapper function, achieving the same results
through application of internal functions and using `igraph` to visualize the
networks at various steps.

## Information retrieval

The first step is to retrieve JSON tree structures of each ontology term. This
is done with the `.getURLs` function, which makes use of the `rols` package.

```{r get_trees}
ncit_ids <- ids[grepl("NCIT", ids)]
ncit_urls <- .getURLs("NCIT", ncit_ids)
ncit_trees <- lapply(ncit_urls, fromJSON)
```

## Network graph creation

These JSON trees can then converted to `igraph`-compatible data structures with the `.createNetwork`, `.groupRoots`, and `.createGraph`
functions, then visualized with `plotNet`.

```{r tree_to_network, fig.cap = "Network diagram of term arrangement within the ontology"}
tree_nets <- lapply(ncit_trees, .createNetwork)
tree_groups <- .groupRoots(tree_nets)
full_net <- bind_rows(tree_groups[[1]])
all_nodes <- unique(unlist(full_net))
original_nodes <- all_nodes[all_nodes %in% ncit_ids]
network_graph <- .createGraph(full_net, original_nodes)

set.seed(1003)
net_layout <- layout_with_fr(network_graph)

plotNet(network_graph, layout = net_layout)
```

## Cutoff filtering

We can now examine the effect of different cutoff values by highlighting nodes
that are designated as "high-traffic." These high-traffic nodes are detected
with the `.busyNodes` function, and highlighted and plotted with `markNet`.

The blue polygon on each graph gives a relative idea of the amount of the graph
that is excluded from candidacy in picking representative nodes, with the
specific nodes highlighted white.

```{r cutoff_filtering, fig.cap = "Nodes designated as \"High-traffic\" based on cutoff value"}
busy0.05 <- .busyNodes(network_graph, length(all_nodes) * 0.05)
busy0.25 <- .busyNodes(network_graph, length(all_nodes) * 0.25)
busy0.95 <- .busyNodes(network_graph, length(all_nodes) * 0.95)

markNet(network_graph, mark_nodes = busy0.05, mark_label = "High-traffic term",
        layout = net_layout)
title(sub = "Cutoff = 0.05")

markNet(network_graph, mark_nodes = busy0.25, mark_label = "High-traffic term",
        layout = net_layout)
title(sub = "Cutoff = 0.25")

markNet(network_graph, mark_nodes = busy0.95, mark_label = "High-traffic term",
        layout = net_layout)
title(sub = "Cutoff = 0.95")
```

## Clustering

After visualizing the high-traffic nodes, we can see how removal of them impacts
igraph's `cluster_fast_greedy` function. This function is used to group ontology
terms into clusters for future representative node selection. We can again
compare the cluster arrangement as it is affected by different cutoff values.

```{r cutoff_clustering, fig.cap  = "Differences in node clustering when \"High-traffic\" nodes are removed at different cutoff values"}
coms0.05 <- communities(cluster_fast_greedy(as.undirected(delete_vertices(
    network_graph, busy0.05))))
coms0.25 <- communities(cluster_fast_greedy(as.undirected(delete_vertices(
    network_graph, busy0.25))))
coms0.95 <- communities(cluster_fast_greedy(as.undirected(delete_vertices(
    network_graph, busy0.95))))

plotClusters(network_graph, coms0.05, layout = net_layout)
title(sub = "Cutoff = 0.05")

plotClusters(network_graph, coms0.25, layout = net_layout)
title(sub = "Cutoff = 0.25")

plotClusters(network_graph, coms0.95, layout = net_layout)
title(sub = "Cutoff = 0.95")
```

## Picking representatives

We then choose representative nodes for each cluster individually. For most
clusters, the `.LCA` function simply chooses the lowest common ancestor of all
of the original terms within the cluster. For some clusters, however, the LCA is
also classified as a high-traffic term and a group of "lower" terms must be
chosen instead. This is done with the function `.bestRoots`, which compares the
in- and out-degrees of each term within the cluster and chooses a group of nodes
that are ancestors of the original terms in the cluster.

```{r LCA, fig.cap = "Picking a representative node for a single cluster by finding the Lowest Common Ancestor"}
picked_nodes <- .LCA(network_graph, coms0.25[[3]])

net <- network_graph
V(net)[picked_nodes]$type <- "picked"
plotClusters(net, list(coms0.25[[3]]), layout = net_layout)
title(sub = "Cutoff = 0.25")
```
```{r best_roots, fig.cap = "Picking a representative node for a single cluster by choosing an optimal group of nodes"}
picked_nodes <- .bestRoots(network_graph, coms0.25[[2]],
                           coms0.25[[2]][coms0.25[[2]] %in% original_nodes])

net <- network_graph
V(net)[picked_nodes]$type <- "picked"
plotClusters(net, list(coms0.25[[2]]), layout = net_layout)
title(sub = "Cutoff = 0.25")
```

---

# Overview of results

Finally, we can see all of the chosen representative nodes for the graph when
different values are used for the cutoff.

```{r picked_cutoffs, fig.cap = "Differences in representative node selection based on cutoff value"}
picked0.05 <- .clusterNodes(network_graph, original_nodes,
                            length(all_nodes) * 0.05)
picked0.25 <- .clusterNodes(network_graph, original_nodes,
                            length(all_nodes) * 0.25)
picked0.95 <- .clusterNodes(network_graph, original_nodes,
                            length(all_nodes) * 0.95)

net0.05 <- network_graph
net0.25 <- network_graph
net0.95 <- network_graph

V(net0.05)[picked0.05]$type <- "picked"
V(net0.25)[picked0.25]$type <- "picked"
V(net0.95)[picked0.95]$type <- "picked"

plotNet(net0.05, layout = net_layout)
title(sub = "Cutoff = 0.05")

plotNet(net0.25, layout = net_layout)
title(sub = "Cutoff = 0.25")

plotNet(net0.95, layout = net_layout)
title(sub = "Cutoff = 0.95")
```